---
title: "607_homework_3"
author: "Alec"
date: "9/9/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries

```{r}
library(tidyverse)
library(stringr)
library(htmlwidgets)
```

# Question 1

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

## Load the data from github using readr package

The readr package is a great way to load data from github

```{r load data}

library(readr)

major_data <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv")

```

## write code that receives the major_dataset, a character vector containing desired major search terms, and returns the matching major names

```{r }

retrieve_majors_with_x <- function(major_data, words) {
  
  pipe_words <- str_c(words, collapse= "|")
  
  return(major_data$Major %>% str_subset(pipe_words))
}

```

## test the formula with "DATA" and "STATISTICS"

```{r}
input_major_string <- c("DATA","STATISTICS")

final_value <- retrieve_majors_with_x(major_data, input_major_string)

final_value
```

## test the formula with anothr set of search terms

```{r}
input_major_string <- c("BUSINESS","DATA","STATISTICS","PSYCHOLOGY")

final_value <- retrieve_majors_with_x(major_data, input_major_string)

final_value

```


# Question 2
Write code that transforms 'input_data' (in the form of a .CSV file) into 'expected_output'

## load the input and expected output data

```{r load input_data and expected_output data}

input_data <- read.csv("homework_2_q_2.csv", header = FALSE)

expected_output <- c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

input_data
```

## write code that converts input to output

My approach is to first concatenate all of the rows of the csv file into one string. Following this I will use regex to remove the [##] values. At this point, I can extract any text that is surrounded by quotation marks. Following this extraction, I remove the quotation marks, and the result will match the expected data.

```{r write code that converts input_format into output_format}
format_conversion <- function(input_data) {
  # first we concatenate all the rows of the .CSV file
  
  concat <- ""
  
  for (x in 1:nrow(input_data)) {
    concat <- str_c(concat, input_data[x,])
  }
  
  # next we remove all '[#]'s from the concatenated string and extract everything where there is text/spaces between quotation marks                 eg. '"apple sauce"'
  
  extracted_values <- concat %>% 
    str_replace_all("\\[\\d+\\]\\s", "") %>%
    str_extract_all("\"[a-z\\s]+\"")
  
  #finally, convert extracted_values into a character vector, and remove all quotation marks
  
  final_format <- str_replace_all(unlist(extracted_values[1]),"\"","")
  
  return(final_format)
}

```

Another, simpler function I made after some more thought:

```{r}
convert_input <- function(input_data) {
  
  ret <- str_c(input_data$V1, collapse="") %>% 
    str_extract_all('[A-Za-z]+.?[A-Za-z]+')
  
  return(unlist(ret[1]))
}
```


## run code, and verify that output == expectd_output for both functions

```{r}
output <- format_conversion(input_data)

output2 <- convert_input(input_data)
```

```{r}
output == expected_output
```

```{r}
output2 == expected_output
```

# Quesiton 3 - Describe what these expressions will match

## (.)\1\1

Without quotations, this should be interpreted as a "numeric expression". As such, this expression would match against any substring that occurs three times in a row within a string.

But if you try to enter this into R with quotations, you will match against something entirely different, and unexpcted. You would match against a substring that contains any character, followed by two escaped 1's:

```{r}
example <- c("brrr\1\1r","hmmm\1","assass\2\2in")
str_extract(example,"(.)\1\1")
```


If you want to get three characters in a row, then you need to convert the numeric expression to a string expression, by escaping the backslashes: "(.)\\1\\1"

The regex "(.)\\1\\1" would match substrings with three repeating characters.

```{r}
example <- c("brrrr","hmmmattt","assassin")
str_extract_all(example,"(.)\\1\\1")
```

## "(.)(.)\\2\\1"

This will match 4 character palindrome substrings

```{r}
example <- c("racecar","woowzers","rocks")
str_extract(example,"(.)(.)\\2\\1")

```

## (..)\1

This is similar to the first one. It will match any two characters followed by an escaped 1 (\1) unless you convert to a string expression.

```{r}
example <- c("hello\1","h\1","bingo")
str_extract(example, "(..)\1")

```
When using double slashes, like "(..)\\1", then it would match a four character string, starting with the first two characters repeating

```{r}
example <- c("banana","titilating","money")
str_extract(example, "(..)\\1")
```

## "(.).\\1.\\1"

This will match a character X, followed by a single character (unspecificed), followed by X, followed by a character (unspecified), followed by X.

```{r}

example <- c("T-T6T Freight Engine", "Monopoly","battleship")
str_extract(example, "(.).\\1.\\1")

```


## "(.)(.)(.).*\\3\\2\\1"

This will match a substring whose last three characters are the reverse of the first three characters. However there must be any number of characters between the first three, and the last three.

```{r}

example <- c("123thisisnotmypassword321","hello_mister_sim","dogs")
str_extract(example, "(.)(.)(.).*\\3\\2\\1")

```

# Question 4 - construct regex to match the following

## Start and end with the same character.

```{r}
pattern <- "(.).*\\1"
example <- c("nylon","velcro","Mr. Greenberg")
str_extract(example, pattern)


```


## Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

```{r}
pattern <- ".*(..).*\\1.*"
example <- c("church","bigbird","other")
str_extract(example, pattern)
```

## Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

```{r}
pattern <- ".*(.).*\\1.*\\1.*"
example <- c("eleven","bananana","mount sinai")
str_extract(example, pattern)
```

pnorm(.6666666666) - pnorm(-.6666666666)

