---
title: "607_homework_9"
author: "Alec"
date: "10/24/2021"
output: 
  prettydoc::html_pretty:
    theme: cosmo
    highlight: github
    keep_md: true
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}
library(jsonlite)
library(tidyverse)
library(stringr)
```


## Connecting to API

Key and secret provided from NYTimes API site: "https://developer.nytimes.com/"

```{r}
key <- "APHySXOpNOGSc10kLBVm8ZdVljlRIRlf"
secret <- "2JYUVWVz9f6QGrkS"
```

With the search api, we can list key-words to filter for, and also filter by time. For this assignment, we will look for articles related to "data-science", published anytime between the start of our semester, to today.

```{r}
term <- "data-science"
begin <- "20210816"
end <- "20211024"
```

We create a "base_url" with all of the required parameters for API querying using stringr::str_c

```{r}
base_url <- str_c("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=", term, "&begin_date=", begin, "&end_date=", end, "&facet_filter=true&api-key=", key, sep="")
```

## Extracting data into dataframe

First we will perform an initial query which will return a set of documents. Part of the information we will receive contains the "maxPages" from the search. Taking this number, we will iterate through values 1:maxPages and extract all documents for each page in the page range. All of the results will be flattened into a dataframe, and then loaded into a list.

```{r}
initialQuery <- fromJSON(base_url)

#maxPages <- round((initialQuery$response$meta$hits[1] / 10)-1) 

pages_2021 <- vector("list",length=15)

for(i in 0:15){
    nytSearch <- fromJSON(paste0(base_url, "&page=", i), flatten = TRUE) %>% data.frame()
    pages_2021[[i+1]] <- nytSearch
    Sys.sleep(5)
}
```

Now that we have all of our dataframes in a list, we can use a simple r_bind call to aggregate them all into the same dataframe

```{r}

nyt_2021_articles <- rbind_pages(pages_2021)

```

Looking below, we have successfully extracted and loaded over 200 NYTimes articles into a dataframe.

```{r}
head(nyt_2021_articles)
```
## Conclusion

Based on the above, it is very straightforward to extract data from the NYTimes search API. In future exploration, I will further examine some of the nested columns that are present (like keywords and persons).


