---
title: "607_homework_2"
author: "Alec"
date: "9/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load RMySQL

RMySQL package allows for consumption of data from MySQL into R. For this assignment I experimented with MySQL, PostgreSQL, and SQLite.

```{r load RMySQL}
library("RMySQL")
```

##Create connection 'mydb'

RMySQL requires a connection to be made to the SQL database. I am using my local connection, with username = "root" and no password setup. In future assignments I will look into building my databases in the cloud and implementing password authentication for tighter security.

```{r connect to SQL database}
mydb = dbConnect(MySQL(), user = 'root', dbname='data_607', host='localhost')
```

## Select data from connection database

Using a simple select statement, we are able to load all of the movie preference data from MySQL into R.

```{r collect SQL data}
data = dbGetQuery(mydb, "select * from data_607.movie_preferences")
```

## Inspect Data

For this assignment I included 8 movies. Most of the films have respondent rankings for all respondents, but there are some exceptions. For instance, no one in my family watched Black Widow.

```{r view data.frame}
data
```

## Dealing with Null Values

There are many methods to dealing with null values. These include marking the nulls with a new character or integer, imputing null values based on surrounding data, or removing columns / rows with excessive null values. 
1. Because Black Widow has no rankings, I will choose to remove this row.
2. For all other NULL values, I will replace with '0', set to mark 'did not watch'

```{r removing black widow row}

data <- data[-c(2), ]

```


```{r replacing remaining NULL values with 0}

data[is.na(data)] = 0

```


```{r view final data.frame}

data

```

