---
title: "Child Mortality Analysis"
author: "Alec"
date: "10/3/2021"
output: 
  prettydoc::html_pretty:
    theme: cosmo
    highlight: github
    keep_md: true
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Discussion Board Title: Global Child Mortality Rates

Dataset: https://sejdemyr.github.io/r-tutorials/basics/data/RatesDeaths_AllIndicators.xlsx

Provided by: Alec McCabe

Suggested Prompt: "What 10 countries have the highest under-5 mortality rates today? For the 10 worst countries, visualize the under-5 mortality trend over time. Comparatively, what does the trend for G7 countries look like over time?

# Load Libraries

For this task, we will be using the tidyverse and stringr libraries

```{r}
library(tidyverse)
library(stringr)
```

# Load Data

Input data has been cleaned (overview and instructions removed from csv) and loaded into a new csv titled child_mortality.csv

```{r}
input_data <- read_csv("child_mortality.csv")
```

### Capture only under-5 mortality rate

This dataset includes information for under-5 mortality rate, infant mortality rate, neonatal mortality rate, and number of deaths. For our analysis, we will be focusing on under-5 mortality.

```{r}
under_5_mortality <- input_data %>%
  rename(iso_code = `ISO Code`, country = CountryName, measurement = `Uncertainty bounds*`) %>%
  select(iso_code, country, measurement, contains("U5MR"))
```

### Use gather() to convert into long format. We will use gather() on the year column to create new key-value columns "year" and "value"

```{r}
under_5_mortality <- under_5_mortality %>%
  gather("year","value",contains("U5MR")) %>%
  mutate(
    year = as.integer(str_replace(year, "U5MR.",""))
  ) %>%
  filter(measurement == "Median") %>%
  select(-measurement)
```

Looking at the data below, we now have a "long" dataframe which includes a single column designating the year for each country observation. Notice below that there are many NA values, presumably because not all countries had tracking begin at the same time. 

```{r}
head(under_5_mortality, n=10)
```

# Answer the Original Prompts:

### Which countries have the highest recentl (2015) rate of under-5 mortality?

Angola, Chad, Somalia, Central African Republic, Seirra Leone, Mali, Nigeria, Benin, Congo DR, Niger had the highest under-5 mortality rates based in 2015. It should be noted that all of these countries are in Africa.

```{r}
under_5_mortality %>%
  filter(year==max(year)) %>%
  arrange(desc(value)) %>%
  head(10)
```
### Visualize the under-5 mortality trend for the 10 worst countries

Collect the names of the countries shown above into a character vector

```{r}
worst_countries <- under_5_mortality %>%
  filter(year==max(year)) %>%
  arrange(desc(value)) %>%
  head(10) %>%
  .$country
```

Create a new dataframe to hold these worst countries, using filter() and %in% to match against the "worst_countries" character vector.

```{r}
under_5_mortality_worst <- under_5_mortality %>%
  filter(
    country %in% worst_countries
  )
```

Despite the fact that these countries have the worst 2015 under-5 mortality rates, it is evident that they have exhibited improvement (dropped rates) over time. Their improvement is roughly linear.

```{r}
under_5_mortality_worst %>%
  ggplot(mapping=aes(x=year, y=value, color=country)) +
  geom_line()
```


### What does the trend for G7 countries look like?

We expect that the under-5 mortality rates for G7 countries to be lower than that of the worst countries, but how will the trend differ?

```{r}
under_5_mortality_g7 <- under_5_mortality %>%
  filter(
    country %in% c("Canada", "United States of America","France","Germany","Italy","Japan","United Kingdom")
  )
```

Based on the below line graph, it is evident that under-5 child mortality rates have been steadily decreasing over the years for G7 countries. Unlike the trend for mortality rates in the "worst countries" group, this downward trend is not linear.

```{r}
under_5_mortality_g7 %>%
  ggplot(mapping=aes(x=year, y=value, color=country)) +
  geom_line()
```

